Testing and Isolation Efficacy; Insights from a Simple Epidemic Model

### 2020-11-18 SIR_manuscript.tex
intro updated

- Non-random testing/testing-tracing-isolation (TTI) is more efficent in controling an epidemic event.
- Potential advantage of slow reporting of test result. 
- Literature reviwe, evidences of efficacy of TTI
- Simple ODE models provide useful insights (some classic examples, Kermack-Makendric,...)
Historically, Ronald Ross (I think) developed a series of mathematical models to study the spread of malaria and implementation of control. His work contributed to the modern theory of the control of
infectious disease (ref: Ross R. The Prevention of Malaria, 2nd edn. London: Murray, 1911).

- Covid-19 and chalanges, questions to be addressed with a simple model.
This study provides a simple deterministic model/tool in which testing mechinisms and isolation have been modeled to study a pandamic in a population.

- What did we do? Why novel? 
(i) We developed a deterministic compartmental model which incorporates testing strategies and isolation mechanism to get insights.
(ii) the potential advantage of slow reporting of test result on the spread of the infection was addressed.
- The basic reproduction number was derived analytically as a threshold of the dynamics. 
- Our results include...

% %%%%%%%
\subsection{Literature Review}

\subsubsection{Explicit models of TTI (trace/test/isolate) based on network or agent-based models}
\citep{endo2020implication} [Ali: It seems to me that this is just a statistical model to estimate the parent-offspring of an infected index, not sure if it fits into agent-based group!] Used simulation on a branching process model to assess the forward and backward contact tracing efficiency. Assuming a negative-binomial branching process with a mean R, reproduction number, and overdispersion parameter k, the mean total number of generation G3 and averted G3 are estimated. The effectiveness of TTI is defined as the ratio of averted to the mean.

\citep{rice2020effect} Effect of school closures on mortality. Reproduce Report 9 results by spatial agent based CovidSim. 
% %%%%%%%
\subsubsection{Models of repeated random testing of isolated populations}
\cite{bergstrom2020frequency}
(1) Model, assumptions: They developed a function, namely expected exposure $E(C,\tau)$, to approximate trade-offs between the frequency of testing, n, the sensitivity of testing, q, and the delay between
testing and results, d. This function is explicitly derived and was connected the effective reproduction number $R=R_0 S$, where $S$ is the proportion of population susceptible.
assumption that transmission rates are a step function: individuals who
have COVID go from non-infectious to fully infectious instantaneously,
and remain fully infectious until they are no longer able to transmit disease. Test sensitivity takes the same form over the course of infection.
More sophisticated models could allow varying infectiousness and varying
sensitivity over time, as in 
\citep{larremore2020test}.

\citep{tuite2020mathematical} used an age-structured compartmental model of COVID-19 transmission in the population of Ontario, Canada. We compared a base case with limited testing, isolation and quarantine to different scenarios. 
% %%%%%%%
\subsubsection{Other maybe-related works}
\citep{arino2020simple} developed a SLIAR compartmental model to study the spread of an epidemic, specifically COVID-19, in a population. The model incorporates an Erlang distribution of times of sojourn in incubating, symptomatically and asymptomatically infectious compartments. Basic reproduction number is derived. Also, sensitivity analysis with respect to the underlying parameters for the following two outputs was carried out; (i) the number of observable cases during the course of the epidemic and at the peak, and (ii) the timing of the peak of the outbreak. Sensitivity analysis is performed using the R package multisensi.

@@

\begin{figure}[h!]
\centering
\begin{subfigure}[t]{.45\textwidth}
\centering
\includegraphics[width=\linewidth]{./pix/R0contour_random.png}
\end{subfigure}
%
\begin{subfigure}[t]{.45\textwidth}
\centering
\includegraphics[width=\linewidth]{./pix/R0contour_TTI.png}
\end{subfigure}
\caption{Behaviour of the basic reproduction number, $\Rnum$, with respect to changes in the underlying parameters; remind the reader about the parameters and their values in this simulations.}
\label{fig:panels}
\end{figure}

@@

>>>>

One of the real chalanges interpreting COVID-19 dynamics is to unravel/investigate information gained from data. One of the strongest singnals we get to see is the daily case reports which are driven by a combination of two processes: (i) epidemic processes and (ii) testing processes. The real chalange here is to seperate these two processes. Simple mathematical models have a long history in providing useful insights facing with real epidemiological chalanges \citep{ross1911prevention} [maybe add influenza work, ideas?]. We developed a model to incorporate mechanistic processes of epidemic processes and testing. We used the model to study the potential effect of testing strategies and slow-fast reporting testing result.

There are different testing strategies with test-specific efficacy, the time required for the results to be reported and the cost of the test. Testing strategies can be grouped into two broad cathegories including (i) random testing across the whole population and (ii) Non-random/bios testing. Note that contact tracing is an example of non-random testing strategies. Each of these strategies has been used as a tool to study COVID-19; see (refs) for group (i) and (refs) for group (ii). The advantage of non-random testing comparing to random testing has been investigated specifically during COVID-19 pandemic (refs). Further, lots of efforts have been invested in developing widely accessible and fast tests in order to metigate the spread of the pandemic (refs). As a comparison of slow versus fast test, the viral PCR test, which is the most accessible test and takes about four days, is very slow relative to the potential tests such as breath-based test \citep{ruszkiewicz2020diagnosis}, which is still in trial stage and takes a few seconds. The point to be made is that there can be a potential advantage of slow test.

In this work we developed a model to incorporate epidemic processes and testing processes mechanisticly to study the potential effect of testing strategies and slow-fast testing. The analysis of the deterministic model included deriving the basic reproduction number, $\Rnum$, as the index of threshold dynamics and the effect of underlying parameters on this key index. We found that the non-random testing strategy is more effective than the random testing. Specifically, the required critical testing intensity to keep $\Rnum$ lower than the threshold value of one, is much lower in the non-random testing than the random one. Further, we show that there is a potential advantage of slow reporting of test result on the spread of the infection.

@@

[comments about the figure:]
1- use a single legend, change the range manually.
\begin{figure}[h!]
\centering
\begin{subfigure}[t]{.45\textwidth}
\centering
\includegraphics[width=\linewidth]{./pix/R0contour_random.png}
\end{subfigure}
%
\begin{subfigure}[t]{.45\textwidth}
\centering
\includegraphics[width=\linewidth]{./pix/R0contour_TTI.png}
\end{subfigure}
\caption{Behaviour of the basic reproduction number, $\Rnum$, with respect to changes in the underlying parameters; remind the reader about the parameters and their values in this simulations.}
\label{fig:panels}
\end{figure}

@@

% %%%%%%%
\section{Literature Review}

\subsection{Explicit models of TTI (trace/test/isolate) based on network or agent-based models}
\citep{endo2020implication} [Ali: It seems to me that this is just a statistical model to estimate the parent-offspring of an infected index, not sure if it fits into agent-based group!] Used simulation on a branching process model to assess the forward and backward contact tracing efficiency. Assuming a negative-binomial branching process with a mean R, reproduction number, and overdispersion parameter k, the mean total number of generation G3 and averted G3 are estimated. The effectiveness of TTI is defined as the ratio of averted to the mean.

\citep{rice2020effect} Effect of school closures on mortality. Reproduce Report 9 results by spatial agent based CovidSim. 
% %%%%%%%
\subsection{Models of repeated random testing of isolated populations}
\cite{bergstrom2020frequency}
(1) Model, assumptions: They developed a function, namely expected exposure $E(C,\tau)$, to approximate trade-offs between the frequency of testing, n, the sensitivity of testing, q, and the delay between
testing and results, d. This function is explicitly derived and was connected the effective reproduction number $R=R_0 S$, where $S$ is the proportion of population susceptible.
assumption that transmission rates are a step function: individuals who
have COVID go from non-infectious to fully infectious instantaneously,
and remain fully infectious until they are no longer able to transmit disease. Test sensitivity takes the same form over the course of infection.
More sophisticated models could allow varying infectiousness and varying
sensitivity over time, as in 
\citep{larremore2020test}.

\citep{tuite2020mathematical} used an age-structured compartmental model of COVID-19 transmission in the population of Ontario, Canada. We compared a base case with limited testing, isolation and quarantine to different scenarios. 
% %%%%%%%
\subsection{Other maybe-related works}
\citep{arino2020simple} developed a SLIAR compartmental model to study the spread of an epidemic, specifically COVID-19, in a population. The model incorporates an Erlang distribution of times of sojourn in incubating, symptomatically and asymptomatically infectious compartments. Basic reproduction number is derived. Also, sensitivity analysis with respect to the underlying parameters for the following two outputs was carried out; (i) the number of observable cases during the course of the epidemic and at the peak, and (ii) the timing of the peak of the outbreak. Sensitivity analysis is performed using the R package multisensi.
